---
title: "PBD sampling"
author: "Richel J.C. Bilderbeek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PBD sampling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates

 * When the PBD package samples as expected
 * When the PBD package samples in unexpected ways
    * `youngest` gives longer branches than `oldest`
    * `random` gives shortest branches than `youngest`
    * `random` gives longer branches than `oldest`
 * A solution to that problem
 
```{r}
library(PBD)
```
 

### Setup

Define a plotting function:

```{r}
# Plot the results of PBD::pbd_sim
# '@param out the results of PBD::pbd_sim
plot <- function(out) {
  # Check input
  testit::assert("igtree.extant" %in% names(out))
  testit::assert("stree_youngest" %in% names(out))
  testit::assert("stree_oldest" %in% names(out))
  testit::assert("stree_random" %in% names(out))
  testit::assert(class(out$igtree.extant) == c("simmap", "phylo"))
  testit::assert(class(out$stree_youngest) == "phylo")
  testit::assert(class(out$stree_oldest) == "phylo")
  testit::assert(class(out$stree_random) == "phylo")

  graphics::par(mfrow = c(1, 4), mar = c(5, 4, 6, 2) + 0.3)
  cols <- stats::setNames(c("gray", "black"), c("i", "g"))
  phytools::plotSimmap(out$igtree.extant, colors = cols, fsize = 2)

  sum_youngest <- sum(out$stree_youngest$edge.length)
  sum_oldest <- sum(out$stree_oldest$edge.length)
  sum_random <- sum(out$stree_random$edge.length)

  ape::plot.phylo(out$stree_youngest, edge.width = 2, font = 1,
    label.offset = 0.1, cex = 2, cex.main = 0.75,
    main = paste("\n\n", "youngest", format(round(sum_youngest, 2), nsmall = 2), "\n\n"))
  ape::add.scale.bar()
  ape::plot.phylo(out$stree_random, edge.width = 2, font = 1,
    label.offset = 0.1, cex = 2, cex.main = 0.75, 
    main = paste("\n\n", "random", format(round(sum_random, 2), nsmall = 2), "\n\n"))
  ape::add.scale.bar()
  ape::plot.phylo(out$stree_oldest, edge.width = 2, font = 1,
    label.offset = 0.1, cex = 2, cex.main = 0.75,
    main = paste("\n\n", "oldest", format(round(sum_oldest, 2), nsmall = 2), "\n\n"))
  ape::add.scale.bar()
  graphics::par(mfrow = c(1, 1))
}
```

Write a general function that searches for examples
 
```{r}
#' Create an example phylogeny
#' @param scr Speciation Completion Rate
#' @param sirg Speciation Initiation Rate of Good Species 
#' @param siri Speciation Initiation Rate of Incipient Species
#' @param sampling sampling method, can be 
#'  'expected', 
#'  'ylto' (youngest longer than oldest)
#'  'rsty' (random shorter than youngest)
#'  'rlto' (random longer than oldest)
create_example <- function(
  scr, 
  sirg, 
  siri, 
  sampling,
  erg = 0.0, 
  eri = 0.0, 
  crown_age = 1,
  min_n_species = 1,
  max_n_species = 10000,
  min_n_subspecies = 2,
  max_n_subspecies = 10000,
  rng_seed = 42
) {
  # Check input
  testit::assert(scr >= 0.0)
  testit::assert(sirg >= 0.0)
  testit::assert(siri >= 0.0)
  testit::assert(sampling %in% c("expected", "ylto", "rsty", "rlto"))
  testit::assert(erg >= 0.0)
  testit::assert(eri >= 0.0)
  testit::assert(crown_age >= 0.0)
  testit::assert(min_n_species >= 1)
  testit::assert(max_n_species >= 1)
  testit::assert(min_n_species <= max_n_species)
  testit::assert(min_n_subspecies >= 1)
  testit::assert(max_n_subspecies >= 1)
  testit::assert(min_n_subspecies <= max_n_subspecies)
  
  set.seed(rng_seed)
  pbd_params <- c(sirg, scr, siri, erg, eri)
  while (TRUE) {  
    out <- PBD::pbd_sim(pars = pbd_params, age = crown_age, soc = 2)
    
    # Count subspecies
    n_subspecies <- length(out$igtree.extant$tip.label)
    if (n_subspecies < min_n_subspecies) next
    if (n_subspecies > max_n_subspecies) next
  
    # Count species
    n_species <- length(out$stree_youngest$tip.label)
    if (n_species < min_n_species) next
    if (n_species > max_n_species) next
  
    # Sum the branch lengths
    sum_youngest <- sum(out$stree_youngest$edge.length)
    sum_oldest <- sum(out$stree_oldest$edge.length)
    sum_random <- sum(out$stree_random$edge.length)

    # Only measure when sampling does give different branch lengths
    if (sum_youngest == sum_oldest) next
    
    if (sampling == "expected") {
      # No unexpected things
      if (sum_random < sum_youngest) next
      if (sum_random > sum_oldest) next
      if (sum_youngest > sum_oldest) next
      testit::assert(sum_youngest < sum_oldest)
      testit::assert(sum_youngest <= sum_random)
      testit::assert(sum_random <= sum_oldest)
    } else if (sampling == "ylto") {
      # Younger Less Than Oldest
      if (sum_youngest < sum_oldest) next
      testit::assert(sum_youngest > sum_oldest)
    } else if (sampling == "rsty") {
      # Random Shorter Than Youngest
      if (sum_youngest > sum_oldest) next
      if (sum_random >= sum_youngest) next
      if (sum_random >= sum_oldest) next
      testit::assert(sum_youngest < sum_oldest)
      testit::assert(sum_random < sum_youngest)
      testit::assert(sum_random < sum_oldest)
    } else if (sampling == "rlto") {
      if (sum_youngest > sum_oldest) next
      if (sum_random <= sum_youngest) next
      if (sum_random <= sum_oldest) next
      testit::assert(sum_youngest < sum_oldest)
      testit::assert(sum_random > sum_youngest)
      testit::assert(sum_random > sum_oldest)
    }
    
    # Found an example!    
    return(out)
  }
}
```

## When the PBD package samples as expected

One expects that:

 * `oldest` gives the longer branch lengths
 * `youngest` gives the shorter branch lengths
 * `random` gives intermediate branch lengths

Here are some examples in which this is the case.

First, the simplest phylogeny possible:

```{r fig.width=7, fig.height=5}
out <- create_example(
  scr = 0.1, sirg = 2, siri = 2, sampling = "expected", 
  rng_seed = 42, max_n_subspecies = 3
)
plot(out)
```

For a phylogeny that is one bit more complex, sampling also works as expected:

```{r fig.width=7, fig.height=5}
out <- create_example(
  scr = 0.1, sirg = 2, siri = 2, sampling = "expected", 
  rng_seed = 44, 
  min_n_subspecies = 4,
  max_n_subspecies = 4
)
plot(out)
```

For an even bigger phylogeny, sampling also works as expected:

```{r fig.width=7, fig.height=5}
out <- create_example(scr = 0.5, sirg = 1, siri = 2, sampling = "expected",
  rng_seed = 51, 
  min_n_subspecies = 5, 
  max_n_subspecies = 10,
  min_n_species = 3
)
plot(out)
```


## When the PBD package samples in unexpected ways

### 1. `youngest` gives longer branches than `oldest`

```{r fig.width=7, fig.height=5}
out <- create_example(scr = 0.5, sirg = 1, siri = 2, sampling = "ylto",
  rng_seed = 54, 
  min_n_subspecies = 4, 
  max_n_subspecies = 4,
  min_n_species = 3
)
plot(out)
```

### 2. `random` gives shortest branches than `youngest`

```{r fig.width=7, fig.height=5}
out <- create_example(scr = 0.5, sirg = 1, siri = 2, sampling = "rsty",
  rng_seed = 55, 
  min_n_subspecies = 4, 
  max_n_subspecies = 7,
  min_n_species = 3
)
plot(out)
```

### 3. `random` gives longer branches than `oldest`

```{r fig.width=7, fig.height=5}
out <- create_example(scr = 0.5, sirg = 1, siri = 2, sampling = "rlto",
  rng_seed = 57, 
  min_n_subspecies = 4, 
  max_n_subspecies = 7,
  min_n_species = 3
)
plot(out)
```
